Certified Symbolic Finite Transducers: Formalization and Applications to String Manipulation

### 2025-07-31 Paper/main.tex
polish paper

We evaluate the performance of our SFT product operation by computing the product of the HTML encoding sanitizer SFT with SFAs constructed from randomly generated Unicode user inputs. This experimental setup reflects realistic usage scenarios where diverse character sets and input patterns are encountered.

@@

Listing \ref{lst-smtlib-code} demonstrates the regular expression variant through an example that replaces numeric sequences with the string "\texttt{NUM}". The constraints are satisfiable because the input string \texttt{a} = "\texttt{2024,2025}" contains a substring "\texttt{2025}" that matches the regular expression \texttt{re.+ (re.range "0" "9")} (equivalent to \texttt{/[0-9]+/}), and replacing this match with "\texttt{NUM}" yields the expected output string \texttt{b} = "\texttt{2024,NUM}". But the left-matching will return UNSAT.
%

@@

  We evaluate CertiStrR using benchmarks from SMT-LIB 2024 \cite{smtlib_benchmarks}, focusing on the QF\_S and QF\_SLIA logic fragments. The benchmarks are divided into two categories based on the replacement operations: (1) string-based replacement (\texttt{str.replace}) and (2) regular expression-based replacement (\texttt{str.replace\_re}). Due to CertiStrR's current front-end limitations in supporting the full SMT-LIB language, we preprocess certain operations. For instance, conjunctive assertions of the form \texttt{(assert (and c1 c2))} are decomposed into separate assertions \texttt{(assert c1)} and \texttt{(assert c2)} when \texttt{c1} and \texttt{c2} are string constraints supported by CertiStrR.

@@

The experimental evaluation was conducted on a laptop with an Apple M4 processor and 24 GB of memory, with a time limit of one minute per test. 
The results show average execution times of 0.27 seconds for \texttt{str.replace\_str} and 0.36 seconds for \texttt{str.replace\_re} operations. Test outcomes were classified into three categories: SAT (satisfiable), UNSAT (unsatisfiable), and Inconclusive.
%
An "Inconclusive" result indicates that the solver cannot determine satisfiability, not due to timeout (all tests are finished in one minute) but rather due to inherent limitations of the forward-propagation algorithm inherited from CertiStr \cite{cpp/KanLRS22}. Specifically, when the string constraints do not satisfy the tree property defined in \cite{cpp/KanLRS22}, the forward-propagation algorithm may be unable to reach a definitive conclusion, even when the variable domains remain non-empty after propagation.

Our performance analysis revealed that while the SFT-based replacement operation modeling is efficient, the primary computational bottleneck stems from automata accumulation during  forward-propagation. Consider the following string constraints:
\begin{align}
x &= x_1\texttt{++}x_2;~x = \texttt{replace}(x_3, p, r); \nonumber \\
x &= x_4;~x = \texttt{replace\_re}(x_5, p_1, r_1); \nonumber
\end{align}
where \texttt{++} denotes string concatenation. The variable $x$ appears multiple times on the left-hand side of the equations, causing the forward-propagation algorithm to accumulate automata representations for all constraints: $x_1\texttt{++}x_2$, \texttt{replace}\newline($x_3, p, r$), $x_4$, and \texttt{replace\_re}($x_5, p_1, r_1$). Each accumulation step requires computing the product of the current automaton with the previous result. Given that the product operation has a worst-case complexity of $O(n^2)$, where $n$ represents the automaton size, this repeated accumulation can lead to state explosion.

>>>>

We evaluate the performance of our SFT product operation by computing the product of the HTML encoding sanitizer SFT with SFAs constructed from randomly generated Unicode user inputs. This experimental setup reflects realistic usage scenarios where diverse character sets and input patterns are encountered. The experimental evaluation was conducted on a laptop with an Apple M4 processor and 24 GB of memory, with a one-minute time limit per test.

@@

Listing \ref{lst-smtlib-code} demonstrates the regular expression variant through an example that replaces numeric sequences with the string "\texttt{NUM}". The constraints are satisfiable because the input string \texttt{a} = "\texttt{2024,2025}" contains a substring "\texttt{2025}" that matches the regular expression \texttt{re.+ (re.range "0" "9")} (equivalent to \texttt{/[0-9]+/}), and replacing this match with "\texttt{NUM}" yields the expected output string \texttt{b} = "\texttt{2024,NUM}". But the left-matching will return \texttt{UNSAT}.
%

@@

We evaluate CertiStrR using benchmarks from SMT-LIB 2024 \cite{smtlib_benchmarks}, focusing on the QF\_S and QF\_SLIA logic fragments. The benchmarks are categorized into three groups based on their pattern matching semantics: (1) general singleton matching, (2) left-most matching, and (3) replace-all matching. Patterns can be either regular expressions or constant strings.

Due to CertiStrR's current frontend limitations in supporting the complete SMT-LIB language specification, we preprocess certain operations. For example, conjunctive assertions of the form \texttt{(assert (and c1 c2))} are decomposed into separate assertions \texttt{(assert c1)} and \texttt{(assert c2)} when both \texttt{c1} and \texttt{c2} are string constraints supported by CertiStrR.

@@

The experimental evaluation was conducted on a laptop with an Apple M4 processor and 24 GB of memory, with a one-minute time limit per test. The results show average execution times of 0.27 seconds for \texttt{replace singleton}, 0.36 seconds for \texttt{left-match}, and 0.36 seconds for \texttt{replace all}. Test outcomes were classified into three categories: SAT (satisfiable), UNSAT (unsatisfiable), and Inconclusive.

An "Inconclusive" result indicates that the solver cannot determine satisfiability, not due to timeout (all tests completed within one minute), but due to the inherent incompleteness of the string solver's algorithm. An inconclusive result can occur when the solver cannot decide whether a string variable can be assigned a unique value to make the constraints satisfiable.

Our performance analysis revealed that while the SFT-based replacement operation modeling is efficient, the primary computational bottleneck stems from automata accumulation during  forward-propagation. Consider the following string constraints:
\begin{align}
x &= x_1\texttt{++}x_2;~x = \texttt{replace}(x_3, p, r); \nonumber ~x = \texttt{replace\_re}(x_5, p_1, r_1); \nonumber
\end{align}
where \texttt{++} denotes string concatenation. The variable $x$ appears multiple times on the left-hand side of the equations, causing the forward-propagation algorithm to accumulate automata representations for all constraints: $x_1\texttt{++}x_2$, \texttt{replace}\newline($x_3, p, r$), $x_4$, and \texttt{replace\_re}($x_5, p_1, r_1$). Each accumulation step requires computing the product of the current automaton with the previous result. Given that the product operation has a worst-case complexity of $O(n^2)$, where $n$ represents the automaton size, this repeated accumulation can lead to state explosion.
